A Comparsion of Traffic Engineering Methods for Application Perceived Metrics

### 2015-07-20 shrink/intro.tex
a finished version: jul 20, 11 AM

The practicality of consolidation as an energy-saving tool for a \cdc\ operator depends on its impact on end-user response time. End-user response time is a key metric on which  operators are evaluated \cite{comparingCDN}. Further, a sharp response time inflation may be perceived as service unavailability, thereby causing a service-level agreement (SLA) violation and revenue loss for an operator \cite{amazonSLA, microsoftSLA, hpSLA}. Despite a large body of literature on consolidation schemes, quantifying the precise impact on user-perceived response times is not well understood today.

Our position is that a lack of quantitative understanding of energy vs. response time tradeoff is potentially a major roadblock to widespread deployment of consolidation to reduce \cdc\ energy use. To provide the insights needed for an operator to reduce its energy use, we address these key questions: (1) What is the energy-response time tradeoff achieved by simple schemes for server and network consolidation in a \cdc\ that uses well-known schemes for load balancing and network routing? (2) What is the additional energy savings achieved by network-aware server consolidation over network-unaware server consolidation considered previously?

Our primary contribution is to quantify the tradeoff between energy savings via consolidation and response times for a real \cdc's workload, and the design and implementation of a system, \shrink, to leverage this tradeoff. \shrink\ reduces the energy use of servers and switches via consolidation, while enabling operators to achieve the desired response time and hardware reliability. A novel aspect of \shrink\ is a network-aware server consolidation scheme that selects the active servers in a left-to-right order in a topology and achieves greater network energy savings over network-unaware server consolidation schemes. Further, our server and network consolidation schemes require only ECMP (equal-cost multipath) support for routing, and hence, are deployable with existing datacenter network fabrics.

Our work is grounded in implementation, due to which it accounts for several factors affecting response time - increased load on servers and network links, reduced storage and its impact on cache hit rates, non-steady state cache behavior due to on-off transitions, imperfect load balancing among servers - and hence, accurately quantifies the impact of consolidation on end-user response time. A key insight, supported via experiments, is that cache hit rates, despite server consolidation, remain close to an unconsolidated datacenter. A small reduction in hit rates helps mitigate the impact of consolidation on response times. This finding is explained by the skewed content popularity obseved in real workloads, due to which working set size of content remains small compared to the storage available in a CDC.

@@

%\shrink\ uses some well-known ideas, namely, randomization for load balancing and ECMP (equal-cost multipath) for routing; their use is in line with our goal of quantifying the energy-response time tradeoff for load balancing and routing schemes that we expect to be widely used in \cdc s.

% -- and hence, accurately quantifies the impact of consolidation on response time due to consolidation.

% to accurately model the relation between server utilization and response time, impact of server consolidation on cache hit rates, and non-ideal load balancing. For these reasons,  our work provides an accurate characterization of energy-response time tradeoffs for \cdc s.

%A challenge \cdc\ operators face is to reduce their energy use with meeting service-level agreements (SLAs) dependent on end-user response times. Energy is a major factor in the operational costs of a datacenter \cite{barroso2007case}  and contributes to 15-20\% of total cost of ownership of a datacenter \cite{GreenbergCost,rasmussen2011determining,power-cost}. A promising approach for reducing datacenter energy is \emph{consolidation}, in which the load on the datacenter is aggregated on a fraction of components, which enables the remaining components to be turned off for saving energy. But, in the context of a \cdc, consolidation can potentially inflate end-user response times by increasing the load on servers and switches that are active and increasing the cache miss rates.  As an inflation in response times could incur violations of a \cdc's SLAs, the practicality of energy saving by consolidation for \cdc s (and other datacenters also) cannot be evaluated without accurately quantifying the impact of those schemes on end-user response times.

%Our work makes two other contributions. First, we propose a network-aware server consolidation scheme that selects the active servers in a left-to-right order in a topology and increases the potential network energy savings.  Second, we present a simple model, which shows that energy vs. response time tradeoff is more favorable as the skew in the server's utilization vs. delay profile increases and the skew in the content popularity distribution increases.

%Second, we present a simple model of energy use and response time in \cdc s, which shows that energy savings can be achieved with a small response time inflation for Zipf content workloads, provided server queuing delay remains a small fraction of the response time even at a high server utilization. Our analysis also shows that the energy-response time tradeoff improves as the skew in the content popularity increases.

@@

(2) \emph{Comparison to ideal:}  \shrink\ achieves a mean, 95-th \%-ile and 99-th \%-ile response time that is 15\%, 3\% and 25\% higher than a scheme that characterizes the ideal energy-response time tradeoff while using 5\% more energy than it.

>>>>

The practicality of consolidation as an energy-saving tool for a \cdc\ operator depends on its impact on user-perceived latency. User-perceived latency is a key metric on which  operators are evaluated \cite{comparingCDN}. Further, a sharp latency inflation may be perceived as service unavailability, thereby causing a service-level agreement (SLA) violation and revenue loss for an operator \cite{amazonSLA, microsoftSLA, hpSLA}. Despite a large body of literature on consolidation schemes, quantifying the precise impact on user-perceived latencies is not well understood today.

Our position is that a lack of quantitative understanding of energy vs. latency tradeoff is potentially a major roadblock to widespread deployment of consolidation to reduce \cdc\ energy use. To provide the insights needed for an operator to reduce its energy use, we address these key questions: (1) What is the energy-latency tradeoff achieved by simple schemes for server and network consolidation in a \cdc\ that uses well-known schemes for load balancing and network routing? (2) What is the additional energy savings achieved by network-aware server consolidation over network-unaware server consolidation considered previously?

Our primary contribution is to quantify the tradeoff between energy savings via consolidation and latencies for a real \cdc's workload, and the design and implementation of a system, \shrink, to leverage this tradeoff. \shrink\ reduces the energy use of servers and switches via consolidation, while enabling operators to achieve the desired latency and hardware reliability. A novel aspect of \shrink\ is a network-aware server consolidation scheme that selects the active servers in a left-to-right order in a topology and achieves greater network energy savings over network-unaware server consolidation schemes. Further, our server and network consolidation schemes require only ECMP (equal-cost multipath) support for routing, and hence, are deployable with existing datacenter network fabrics.

Our work is grounded in implementation, due to which it accounts for several factors affecting latency - increased load on servers and network links, reduced storage and its impact on cache hit rates, non-steady state cache behavior due to on-off transitions, imperfect load balancing among servers - and hence, accurately quantifies the impact of consolidation on user-perceived latency. A key insight, supported via experiments, is that cache hit rates, despite server consolidation, remain close to an unconsolidated datacenter. A small reduction in hit rates helps mitigate the impact of consolidation on latencies. This finding is explained by the skewed content popularity obseved in real workloads, due to which working set size of content remains small compared to the storage available in a CDC.

@@

%\shrink\ uses some well-known ideas, namely, randomization for load balancing and ECMP (equal-cost multipath) for routing; their use is in line with our goal of quantifying the energy-latency tradeoff for load balancing and routing schemes that we expect to be widely used in \cdc s.

% -- and hence, accurately quantifies the impact of consolidation on latency due to consolidation.

% to accurately model the relation between server utilization and latency, impact of server consolidation on cache hit rates, and non-ideal load balancing. For these reasons,  our work provides an accurate characterization of energy-latency tradeoffs for \cdc s.

%A challenge \cdc\ operators face is to reduce their energy use with meeting service-level agreements (SLAs) dependent on end-user latencies. Energy is a major factor in the operational costs of a datacenter \cite{barroso2007case}  and contributes to 15-20\% of total cost of ownership of a datacenter \cite{GreenbergCost,rasmussen2011determining,power-cost}. A promising approach for reducing datacenter energy is \emph{consolidation}, in which the load on the datacenter is aggregated on a fraction of components, which enables the remaining components to be turned off for saving energy. But, in the context of a \cdc, consolidation can potentially inflate end-user latencies by increasing the load on servers and switches that are active and increasing the cache miss rates.  As an inflation in latencies could incur violations of a \cdc's SLAs, the practicality of energy saving by consolidation for \cdc s (and other datacenters also) cannot be evaluated without accurately quantifying the impact of those schemes on end-user latencies.

%Our work makes two other contributions. First, we propose a network-aware server consolidation scheme that selects the active servers in a left-to-right order in a topology and increases the potential network energy savings.  Second, we present a simple model, which shows that energy vs. latency tradeoff is more favorable as the skew in the server's utilization vs. delay profile increases and the skew in the content popularity distribution increases.

%Second, we present a simple model of energy use and latency in \cdc s, which shows that energy savings can be achieved with a small latency inflation for Zipf content workloads, provided server queuing delay remains a small fraction of the latency even at a high server utilization. Our analysis also shows that the energy-latency tradeoff improves as the skew in the content popularity increases.

@@

(2) \emph{Comparison to ideal:}  \shrink\ achieves a mean, 95-th \%-ile and 99-th \%-ile latency that is 15\%, 3\% and 25\% higher than a scheme that characterizes the ideal energy-latency tradeoff while using 5\% more energy than it.

### 2015-07-20 shrink/related.tex
a finished version: jul 20, 11 AM

\textbf{Impact on web-page load time:} Our prototype-based experiments evaluate the response time for individual HTTP requests, and hence do not capture a key metric that is more relevant from an end-user's perspective: web-page load time. However, we expect the inflation in web-page load time to be lower than the inflation in response times given that computation in web browsers constitutes up to 35\% of the critical path of a web-page load time \cite{wprof}.

@@

Our effort distinguishes from prior work in quantifying the energy-response time tradeoff  in \cdc s, presenting the design and implementation of a system to leverage this tradeoff and proposing a network-aware server consolidation scheme to reduce network energy use.  Prior work on reducing energy of datacenters can be divided into three topics: (1) power-proportionality of servers and switches.  (2) server and network consolidation in a datacenter and (3) global load balancing across datacenters.

@@

\textbf{Server consolidation:} Given the long line of work in server consolidation, our work does not focus on saving more energy than the existing consolidation schemes, but instead on accurately quantifying the impact on response times of a simple consolidation scheme.
%Both analytical and experimental studies of server consolidation have been previously explored.

@@

The analytical work in this area shares similar goals as us.  Lin et al. \cite{lin12} propose an algorithm for optimizing a cost metric that incorporates energy costs, on-off switching costs and cost of degradation in performance. Mathew et al. \cite{mathew12} propose an algorithm that balances energy use, reliability and availability of servers, which they evaluate based on load traces from Akamai datacenters. In comparison, our implementation-based approach enables us to accurately model the relation between server utilization and response time, impact of server consolidation on cache hit rates, and non-ideal load balancing, to accurately quantify the impact of consolidation on response time for \cdc s.

Several efforts have conducted an implementation-based evaluation of server consolidation for stateless systems. Chase et al. \cite{chase2001managing} allocate resources among multiple co-hosted services in a cluster while reducing energy via consolidation. Pinheiro's \cite{pinheiro2001load} system proposes consolidation and load balancing algorithms given a bound on the performance degradation that is acceptable.  Rajamani et al. \cite{rajamani2003evaluating} evaluate consolidation schemes for a modified TPC-W workload. In comparison, our effort focuses on \cdc s that maintain a large amount of state in the form of cached content. In \cdc s, the effect of consolidation on response times cannot be evaluated accurately without accounting for the effect of consolidation on the availability of cached content and the resulting cache hit rates.
%One of our key findings is that consolidation results in a small impact on cache hit rates, which helps \shrink\ achieve a good energy-response time tradeoff.

Trushkowsky et al. \cite{Trushkowsky:2011}  dynamically allocate servers and reconfigure the data stored on the servers to meet service-level objectives such as 99-th percentile request latency.  However, there are two key differences between their work and ours. First, they focus  on a workload exclusively of small (256B) objects stored in-memory, whereas  \cdc s need to deal with orders of magnitude of heterogeneity in object sizes and extensively use a disk cache to improve hit rates.  Second, their system appears to be a backend data store, which always has content available within the datacenter. In comparison, \cdc s have a significant fraction of traffic to remote datacenters due to cache misses, and the impact of consolidation on response time in \cdc s depends on the increase in traffic to remote datacenters that consolidation causes. For these reasons, it is not clear if their findings on the impact of dynamic server allocation on request latency would be applicable for \cdc s.
%One of our key findings is that consolidation results in a small increase in cache miss rates, which helps \shrink\ achieve a good energy-response time tradeoff.

@@

\textbf{Network consolidation:} Network  consolidation has been studied for both wide-area and data center networks \cite{response, elasticTree, greenTE, Chiaraviglio, Andrews}. Network consolidation concentrates traffic, represented in the form of a traffic matrix, on a subset of links and switches, and turns off remaining switches and links to save energy. Our work differentiates from prior work in two ways. First, prior work evaluates schemes mostly using traffic engineering metrics such as link utilization, while we evaluate actual end user response time for a real application and show that network consolidation can be performed with a small performance impact in \cdc s. Second, we show network consolidation is closely related to server consolidation. Our network-aware server consolidation saves up to 45\% more network energy over a network-unaware server consolidation scheme.
%network-aware server consolidation increase the potential savings that a network consolidation scheme can achieve. 
%
%
%
%
%
%
% with little effort on measuring response times for real applications
%
%While previous work has studied server and network consolidation as independent problems, we show that these problems are closely related. For the same number of servers, which set of servers is chosen affects potential energy savings that a network consolidation scheme can achieve. Further, we propose a simple network-aware server consolidation scheme saves up to 45\% more network energy over a network-unaware server consolidation scheme.

>>>>

\textbf{Impact on web-page load time:} Our prototype-based experiments evaluate the latency for individual HTTP requests, and hence do not capture a key metric that is more relevant from an end-user's perspective: web-page load time. However, we expect the inflation in web-page load time to be lower than the inflation in latencies given that computation in web browsers constitutes up to 35\% of the critical path of a web-page load time \cite{wprof}.

@@

Our effort distinguishes from prior work in quantifying the energy-latency tradeoff  in \cdc s, presenting the design and implementation of a system to leverage this tradeoff and proposing a network-aware server consolidation scheme to reduce network energy use.  Prior work on reducing energy of datacenters can be divided into three topics: (1) power-proportionality of servers and switches.  (2) server and network consolidation in a datacenter and (3) global load balancing across datacenters.

@@

\textbf{Server consolidation:} Given the long line of work in server consolidation, our work does not focus on saving more energy than the existing consolidation schemes, but instead on accurately quantifying the impact on latencies of a simple consolidation scheme.
%Both analytical and experimental studies of server consolidation have been previously explored.

@@

The analytical work in this area shares similar goals as us.  Lin et al. \cite{lin12} propose an algorithm for optimizing a cost metric that incorporates energy costs, on-off switching costs and cost of degradation in performance. Mathew et al. \cite{mathew12} propose an algorithm that balances energy use, reliability and availability of servers, which they evaluate based on load traces from Akamai datacenters. In comparison, our implementation-based approach enables us to accurately model the relation between server utilization and latency, impact of server consolidation on cache hit rates, and non-ideal load balancing, to accurately quantify the impact of consolidation on latency for \cdc s.

Several efforts have conducted an implementation-based evaluation of server consolidation for stateless systems. Chase et al. \cite{chase2001managing} allocate resources among multiple co-hosted services in a cluster while reducing energy via consolidation. Pinheiro's \cite{pinheiro2001load} system proposes consolidation and load balancing algorithms given a bound on the performance degradation that is acceptable.  Rajamani et al. \cite{rajamani2003evaluating} evaluate consolidation schemes for a modified TPC-W workload. In comparison, our effort focuses on \cdc s that maintain a large amount of state in the form of cached content. In \cdc s, the effect of consolidation on latencies cannot be evaluated accurately without accounting for the effect of consolidation on the availability of cached content and the resulting cache hit rates.
%One of our key findings is that consolidation results in a small impact on cache hit rates, which helps \shrink\ achieve a good energy-latency tradeoff.

Trushkowsky et al. \cite{Trushkowsky:2011}  dynamically allocate servers and reconfigure the data stored on the servers to meet service-level objectives such as 99-th percentile request latency.  However, there are two key differences between their work and ours. First, they focus  on a workload exclusively of small (256B) objects stored in-memory, whereas  \cdc s need to deal with orders of magnitude of heterogeneity in object sizes and extensively use a disk cache to improve hit rates.  Second, their system appears to be a backend data store, which always has content available within the datacenter. In comparison, \cdc s have a significant fraction of traffic to remote datacenters due to cache misses, and the impact of consolidation on latency in \cdc s depends on the increase in traffic to remote datacenters that consolidation causes. For these reasons, it is not clear if their findings on the impact of dynamic server allocation on request latency would be applicable for \cdc s.
%One of our key findings is that consolidation results in a small increase in cache miss rates, which helps \shrink\ achieve a good energy-latency tradeoff.

@@

\textbf{Network consolidation:} Network  consolidation has been studied for both wide-area and data center networks \cite{response, elasticTree, greenTE, Chiaraviglio, Andrews}. Network consolidation concentrates traffic, represented in the form of a traffic matrix, on a subset of links and switches, and turns off remaining switches and links to save energy. Our work differentiates from prior work in two ways. First, prior work evaluates schemes mostly using traffic engineering metrics such as link utilization, while we evaluate actual end user latency for a real application and show that network consolidation can be performed with a small performance impact in \cdc s. Second, we show network consolidation is closely related to server consolidation. Our network-aware server consolidation saves up to 45\% more network energy over a network-unaware server consolidation scheme.
%network-aware server consolidation increase the potential savings that a network consolidation scheme can achieve. 
%
%
%
%
%
%
% with little effort on measuring latencies for real applications
%
%While previous work has studied server and network consolidation as independent problems, we show that these problems are closely related. For the same number of servers, which set of servers is chosen affects potential energy savings that a network consolidation scheme can achieve. Further, we propose a simple network-aware server consolidation scheme saves up to 45\% more network energy over a network-unaware server consolidation scheme.

### 2015-07-20 tepaper/section1.tex
a finished version: jul 20, 11 AM

We show following results from comparing SPF achieved by TE schemes. The optimal TE scheme with perfect knowledge of traffic matrix and sub-optimal TE schemes like demand-aware TE by optimizing OSPF link weights \cite{fortz2000internet} end-up achieving the same SPF. Even the demand-oblivious scheme, \invcap, achieves an SPF at most 30\% worse than optimal TE. Overall, these results show that application adaptation to location diversity blurs differences between traffic engineering schemes.

The rest of the chapter is as follows. Section \ref{sec:locdiv-background} introduces location diversity and proposes a new capacity metric SPF to compare traffic engineering schemes. Section \ref{sec:exp_setup} presents our simulation setup. Section \ref{sec:app_performance} compares the application performance of TE schemes and Section \ref{sec:capacity} compares their achieved capacity under location diversity.

>>>>

\tbd{fix this} This observation has also been confirmed by explicit measurements on Internet backbones \cite{ExpRouterBuffer}, and is consistent with studies on ISP backbones showing that over 90\% of all packet loss is caused by interdomain routing fluctuations as opposed to high utilization \cite{SprintStudy} and 90\% of TCP flows experience no packet loss \cite{SprintBackbone}. Furthermore, end-to-end Internet path delays are known to be largely determined by propagation delays as opposed to queueing delays \cite{SprintBackbone,SingleHopDelay}.

We show following results from comparing SPF achieved by TE schemes. The optimal TE scheme with perfect knowledge of traffic matrix and sub-optimal TE schemes like demand-aware TE by optimizing OSPF link weights \cite{fortz2000internet} end-up achieving the same SPF. Even the demand-oblivious scheme, \invcap, achieves an SPF at most 30\% worse than optimal TE. Overall, these results show that application adaptation to location diversity blurs differences between TE schemes.

The rest of the chapter is organized as follows. Section \ref{sec:locdiv-background} introduces location diversity and proposes a new capacity metric SPF to compare traffic engineering schemes. Section \ref{sec:exp_setup} presents our simulation setup. Section \ref{sec:app_performance} compares the application performance of TE schemes and Section \ref{sec:capacity} compares their achieved capacity under location diversity.

### 2015-07-20 tepaper/section3.tex
a finished version: jul 20, 11 AM

\subsection{Traffic engineering schemes} We select a subset of TE schemes reflecting a variety of proposed approaches in the literature.

>>>>

\subsection{Traffic engineering schemes} We select a subset of TE schemes reflecting a variety of proposed approaches in the literature including optimal TE (\opt), demand-oblivious TE (\invcap) and demand-aware TE (\optwt, \mplsavg\ and \cope).

### 2015-07-20 tepaper/section8.tex
a finished version: jul 20, 11 AM

Our comparison of TE schemes based on user-perceived metrics while accounting for application adaption to location diversity reveals unexpected results that challenge conventional wisdom in traffic engineering. 
We find that link utilization, the most widely used metric to evaluate TE, is a poor predictor of user-perceived performance. 
Under typical Internet load conditions, all TE schemes and even static routing achieve nearly identical user-perceived performance despite achieving vastly different MLUs.  In fact, engineering link utilization in order to accommodate unexpected traffic spikes  can actually hurt common-case application performance. 
More intriguingly, we find that application adaptation to location diversity, or the ability to download content from multiple locations, eliminates differences in the achieved capacity of all TE schemes including ``optimal" TE. 
With location diversity, even static routing achieves a capacity that is at most 30\% (and typically significantly less) worse than the optimal. Taken together, our findings suggest that it matters little which TE scheme is used at today's traffic levels as well as under reasonable projections of increased demand.

>>>>

Our comparison of TE schemes based on user-perceived metrics while accounting for application adaption to location diversity reveals unexpected results that challenge conventional wisdom in traffic engineering. 
We find that link utilization, the most widely used metric to evaluate TE, is a poor predictor of user-perceived performance. 
Under typical Internet load conditions, all TE schemes and even demand-oblivious static routing achieve nearly identical user-perceived performance despite achieving vastly different MLUs.  In fact, engineering link utilization in order to accommodate unexpected traffic spikes  can actually hurt common-case application performance. 
More intriguingly, we find that application adaptation to location diversity, or the ability to download content from multiple locations, eliminates differences in the achieved capacity of all TE schemes including ``optimal" TE. With location diversity, even demand-oblivious static routing achieves a capacity that is at most 30\% (and typically significantly less) worse than the optimal. Taken together, our findings suggest that it matters little which TE scheme is used at today's traffic levels as well as under reasonable projections of increased demand.
